There are so many people to thank, I cannot hope to list them all here.

None of my efforts before or during graduate school would have been possible without unwavering support from my family.
In then ten years and five different states since college, time and distance have only heightened my appreciation for everything you have given me.
Thank you.
% Thanks also for an ample supply of fish.

Those who aided my journey getting here, especially...
Professor Michael Crowley at the University of Southern California, for his letter of recommendation five years after I graduated.
My coworkers from Lockheed Martin, especially Gerald Nostrand for his letter of recommendation, and for just being an awesome technical manager---thanks for letting us be engineers.

Professor Hank Hoffmann, the best advisor a graduate student could ask for...
Besides the exceptional mentoring and support which would require a whole other document to describe, thanks for off-topic conversations, and for taking me into your confidence far more than most advisors ever would.
And for sending me \emph{everywhere}, especially Portugal, Hong Kong, Greece, and France.
I wish you could have joined in the adventures, but thanks also for trusting me to ``spread the gospel'' and defend our work on my own.
Now that you have tenure, my work here is done!

Also at the University of Chicago, especially...
Professors Andrew Chien and Haryadi Gunawi, for being on my Master's committee.
Professor Shan Lu, for being on my doctoral dissertation committee.
Nikita Mishra and David Kim, for collaborations and co-authorships.
Huazhe (Harper) Zhang and Kevin Zhao, for helping with the grunt work.
Ivana Marin{\u c}i{\'c}, for being well-connected and helping my with my job hunt.
The department technical and administrative staff, especially Bob Bartlett, Margaret Jaffey, Nita Yack, and Sandy Quarles---nothing would ever get done with you, you are all great!

Those who supported my research outside the university, especially...
Lars Bergstrom, for the summer position at Mozilla Research.
Steve Hofmeyr, for research collaboration, hosting me at Lawrence Berkeley National Laboratory, and being on my dissertation committee.
Our Proteus project collaborators at Rice, UT Austin, and MIT, especially Professors Krishna Palem and Don Fussell.

Finally, those providing support beyond research, especially...
Saeid Barati, for long chats, and for finding volleyball and softball groups where we can unwind, and sometimes even win.
Rhys Povey (Physics), for organizing so many social events, barbecues, and our intramural volleyball team.
Neville Eclov (Medical Physics), for having us on the Ioniziers softball team and hosting classy parties in the off-season.
Will Kong, Joe Wingerter, Kavon Farvardin, and Saeid Barati (again), for all the homebrewing and other social activities.
Roselyne Tchoua, for sharing in first-year struggles---it was nice having another adult to talk with!
G{\"o}kalp Demirci, for casual drinks, and being the only one to show up for my first-year Thanksgiving potluck.
Anne Farrell, for cookies, a bicycle, and impressively silent socialization.
All the PSD Happy Hour regulars from the department, for banter over cheap pizza and beverages.

For all members of Hank's group, past and present, especially those not yet mentioned but equally important to me (Zhixuan Zhou, Bernard Dickens, Muhammad Husni Santriaji), thanks for putting up with the sarcastic old guy in the room.
I hope I managed to drop a little wisdom along the way---if you find it, I might need it back.
