\chapter{POET: The Performance with Optimal Energy Toolkit}
\label{sec:poet}

This chapter describes POET (the \textbf{P}erformance with \textbf{O}ptimal \textbf{E}nergy \textbf{T}oolkit).
\secref{related} discussed existing work in managing performance and power/energy awareness.
Like POET, a number of these approaches use feedback control to manage timing constraints \cite{Agilos,Bertini2007,grace2,CoAdapt,Maggio13,TCST,ControlWare,PTRADE,Fu2012,Li2007}.
POET is most related to prior approaches that abstract resource management into a middleware or runtime, like \cite{ControlWare,Sojka,PTRADE,Rajkumar}.
Control techniques provide a formal framework for reasoning about the dynamic behavior of the system.
POET is unique in its energy awareness, in its design for portability, and the incorporation of a true minimal-energy resource allocation algorithm.

\TODO{Combine RTAS and MCSoC papers, including motivation, usage, and evaluation.}

\input{poet-motivation}
\input{poet-framework}
\input{poet-implementation}
\input{poet-usage}
\input{poet-evaluation}


\section{Inspired Projects}
\label{sec:poet-inspired}

\TODO{Find appropriate place for this material, \eg an Appendix?}

POET has also been used as the foundation for other controllers and projects.
We expanded on POET to create Bard, which adds the ability to instead meet soft power constraints and maximize performance \cite{Bard}.
Farrell and Hoffmann build on POET by additionally changing application accuracy to provide hard real-time guarantees \cite{meantime}.
Mishra \etal combine POET with machine learning in a project called CALOREE to reduce the amount of offline work needed to generate resource specifications, and update both the specification and pole value online as new application behavior is learned \cite{CALOREE}.
POET was also the starting point in an ongoing project called Proteus, in which we are developing a programming language called FAST that allows programmers to specify more general constraints and optimizations called \emph{intents}, which uses both system and application knobs to satisfy.

Additionally, POET (and the work that motivated it \cite{Imes2014}) ultimately led to developing EnergyMon, a portable software interface for accessing energy metrics at runtime \cite{energymon}.
EnergyMon was integrated with Mozilla's Servo web browsing engine \cite{servo}.
It is also utilized in other research projects, like the aforementioned FAST language, and in the next project in this thesis, CoPPer.
